\documentclass{article}
\usepackage{booktabs, graphicx, hyperref, fontspec}
\usepackage{sectsty}
\allsectionsfont{\sffamily}
\usepackage[margin=1in]{geometry}
\hypersetup{
  colorlinks = true,
  urlcolor = cyan,
 }
 \providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{Fira Sans}

\begin{document}

\sffamily

\centerline{\Huge Microeconomic Analysis}

\vspace{3 mm}

\centerline{\large Dr.~Ryan Safner}
\vspace{2 mm}
\centerline{\large \href{http://micro.classes.ryansafner.com}{microf19.classes.ryansafner.com}}

\vspace{5 mm}

\begin{tabular}{@{}p{3.5in}p{3.5in}}           
\textbf{Course}: ECON 306 Section 2 & \textbf{Email:}  \href{mailto:safner@hood.edu}{\nolinkurl{safner@hood.edu}}\\
\textbf{Room}: Rosenstock 215 & \textbf{Office:}  Rosenstock 118\\
\textbf{Meets}: MW 2:00--3:15 P.M. & \textbf{Hours:} TuTh 10:30--11:30 A.M.\\ 
\end{tabular}

\vspace{5 mm}

\hrule


\begin{quote}
``The curious task of economics is to demonstrate to men how little they
really know about what they imagine they can design.'' - F.A. Hayek,
1974 Economics Nobel Laureate
\end{quote}

\textbf{Microeconomic Analysis} {[}\^{}courseinfo{]} is a collection of
analytical tools for modeling and understanding how decisions made by
agents (individuals and firms) in a context of market institutions lead
to social coordination and the creation of value. This toolkit is the
analytical foundation for exploring \emph{all} fields and applications
of economics as a social science, has many practical applications in
business, finance, politics, and insights into any realm of human
endeavors. Economics, after all, is the study of human action. At the
intermediate level, these concepts are made more rigorous through the
use of mathematical models, but we will first and foremost lay the
theoretical foundations with economic intuition. As such, this class
assumes you have met the \textbf{required prerequisite courses} by
taking introductory economics\footnote{ECON 205, 206} and basic
mathematics\footnote{MATH 112}.

\hypertarget{course-objectives}{%
\section{Course objectives}\label{course-objectives}}

This class will add rigor to your thinking about the world by empowering
you with the ability to model the choices and interactions of people and
organizations in their economic capacities -- as consumers, producers,
firms, voters, etc.

\textbf{By the end of this course,} you will:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  apply the models of microeconomics (constrained optimization and
  equilibrium) towards explaining real world behavior of individuals,
  firms, and governments
\item
  explore the effects of economic and political processes on market
  performance (competition, market prices, profits and losses, property
  rights, entrepreneurship, market power, market failures, public
  policy, government failures)
\item
  apply the economic way of thinking to real world issues in writing
\end{enumerate}

Given these objectives, this course fulfills all three of the learning
outcomes for
\href{https://www.hood.edu/academics/departments/george-b-delaplaine-jr-school-business/student-learning-outcomes}{the
George B. Delaplaine, Jr.~School of Business} Economics B.A. program:

\begin{itemize}
\tightlist
\item
  Use quantitative tools and techniques in the preparation,
  interpretation, analysis and presentation of data and information for
  problem solving and decision making {[}\ldots{]}
\item
  Apply economic reasoning and models to understand and analyze problems
  of public policy {[}\ldots{]}
\item
  Demonstrate effective oral and written communications skills for
  personal and professional success{[}\ldots{]}
\end{itemize}

\textbf{Fair warning:} \textbf{Economics is hard.} \emph{This, in
particular, may be of the hardest courses that you will take, primarily
due to the mathematical content.} Using the economic way of thinking is
a skill, it is literally retraining your brain to interpret and analyze
the world in a novel way, and is not something that can be memorized. I
will do my best to make this class intuitive and helpful, if not
interesting. If at any point you find yourself struggling in this course
for any reason, please come see me. Do not suffer in silence! Coming to
see me for help does not diminish my view of you, in fact I will hold
you in \emph{higher} regard for understanding your own needs and taking
charge of your own learning. There are also a some fantastic resources
on campus, such as the
\href{http://www.hood.edu/campus-services/academic-services/index.html}{Center
for Academic Achievement and Retention (CAAR)} and the
\href{http://www.hood.edu/library/}{Beneficial-Hodson Library}.

See my \href{/reference\#tips}{tips for success in this course}.

\hypertarget{course-materials}{%
\section{Course materials}\label{course-materials}}

You can find all course materials at my academic website dedicated for
(all sections of) this course
\href{http://microf19.classes.ryansafner.com}{microf19.classes.ryansafner.com}.

My lecture slides will be shared with you, and serve as your primary
resource, but our main ``textbook'' below is \textbf{recommended} as the
next best resource and will be available from the campus bookstore.

\hypertarget{books}{%
\subsection{Books}\label{books}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Goolsbee, Austan, Steven Levitt, and Chad Syverson, 2012,
  \emph{Microeconomics} 2nd ed, USA: Worth Publishers, ISBN:
  978-1464187025
\end{enumerate}

You are welcome to purchase the book by other means (e.g.~Amazon,
half.com, etc). I have no financial stake in requiring you to purchase
this book. You are welcome to use previous version of the book, but
carefully verify the reading assignments, as the material may be
different across versions.

\hypertarget{assignments-and-grades}{%
\section{Assignments and grades}\label{assignments-and-grades}}

\begin{center}

\begin{tabular}{lll}
\toprule
 & Assignment & Percent\\
\midrule
1 & Opinion-Editorial & 20\%\\
n & Homeworks (Average) & 20\%\\
3 & Exams & 20\% each\\
\bottomrule
\end{tabular}
\end{center}

All grades are based on the following traditional scale:

\begin{center}

\begin{tabular}{llll}
\toprule
Grade & Range & Grade1 & Range1\\
\midrule
A & 93–100\% & C & 73–76\%\\
A− & 90–92\% & C− & 70–72\%\\
B+ & 87–89\% & D+ & 67–69\%\\
B & 83–86\% & D & 63–66\%\\
B− & 80–82\% & D− & 60–62\%\\
\addlinespace
C+ & 77–79\% & F & < 60\%\\
\bottomrule
\end{tabular}
\end{center}

These grades are firm cutoffs, but I do of course round upwards
(\(\geq 0.5\)) for final grades. A necessary reminder, as an academic, I
am not in the business of \emph{giving} out grades, I merely report the
grade that you \emph{earn}. I will not alter your grade unless you
provide a reasonable argument that I am in error (which does happen from
time to time).

\hypertarget{policies-and-expectations}{%
\section{Policies and Expectations}\label{policies-and-expectations}}

This syllabus is a contract between you, the student, and me, your
instructor. It has been carefully and deliberately thought out\footnote{A
  syllabus can and will be used as a legal document for disputes tried
  at a court of law. Ask me how I know.}, and I will uphold my end of
the agreement and expect you to uphold yours.

In the language of game theory, this syllabus is my commitment device. I
am a very understanding person, and I know that exceptions to rules
often need to be made for students. However, to be \emph{fair} to
\emph{all} students the syllabus artificially constrains my ability to
make exceptions at a whim for anyone. This prevents clever students from
exploiting my congenial personality at everyone else's expense. Please
read and familiarize yourself with the course policies and expectations
of you. Chances are, if you have a question, it is answered herein.

\hypertarget{attendance-and-participation}{%
\subsubsection{Attendance and
Participation}\label{attendance-and-participation}}

I expect you to attend class and to come having already done the reading
assigned for that day. I will remind you in class and possibly through
Blackboard or email which readings I want you to read for the next
class. You are all adults and I will treat you as such. I do not take
attendance, nor do I grade formally for participation but I strongly
recommended you attend class and participate for your sake and the sake
of your classmates. If you are too distracted or are not prepared to
learn, I suggest you stay home, where you can check Facebook more
efficiently. I reserve the right to boost the final grades of students
that I believe have made consistent, quality contributions above and
beyond their peers in class conversations by up to 2.5 points.

\hypertarget{devices}{%
\subsubsection{Devices}\label{devices}}

You are allowed to have and use laptops and tablets in the classroom. I
will not stop you, but I strongly discourage you from using these to
take notes (see Tips for Success). As a courtesy to myself and to
others, do not use your phone in class. I reserve the right to embarrass
you in front of everyone if you do so.

\hypertarget{absences-and-make-ups}{%
\subsubsection{Absences and Make-Ups}\label{absences-and-make-ups}}

You generally do \emph{not} need to let me know if you are unable to
make class, \emph{unless} it is on the day of an exam. It will however,
be your responsibility to acquire the notes from a classmate for any
missed classes. If you are unable to attend an exam for a legitimate
reason (e.g.~sports/club events, traveling, illness, family issues),
please notify me at least \emph{one week} in advance, and we will
schedule a make-up exam date. If you are ill or otherwise unable to
attend on the day of the exam, contact me ASAP to make arrangements.
Failure to do so, including desperate attempts to make arrangements only
\emph{after} the absence will result in a grade of 0 and little
sympathy. I reserve the right to re-weight other assignments for
students who I believe are legitimately unable to complete a particular
assignment.

\hypertarget{late-assignments}{%
\subsubsection{Late Assignments}\label{late-assignments}}

I will accept late assignments, but will subtract a specified amount of
points as a penalty. See individual assignment descriptions for the
amount of points taken off (as it varies by assignment). If an answer
key is posted before you turn in your assignment, the maximum grade you
can earn is an 80. Even if it is the last week of the semester, I
encourage you to turn in late work: some points are better than no
points!

\hypertarget{grading}{%
\subsubsection{Grading}\label{grading}}

I will try my best to post grades on Blackboard's Grading Center and
return graded assignments to you within about one week of you turning
them in. There will be exceptions. Where applicable, I will post answer
keys once I know most homeworks are turned in (see Late Assignments
above for penalties). Blackboard's Grading Center is the place to look
for your most up-to-date grades. You will also be given an Excel
spreadsheet template where you can calculate your overall grade and
forecast ``what if'' scenarios.

\hypertarget{email-accounts}{%
\subsubsection{Email Accounts}\label{email-accounts}}

Students must regularly monitor their Hood email accounts to receive
important college information, including messages related to this class.
Email through the Blackboard system is my main method of communicating
announcements and deadlines regarding your assignments. \textbf{Please
do not reply to any automated Blackboard emails - I may not recieve
it!}. My Hood email
(\href{mailto:safner@hood.edu}{\nolinkurl{safner@hood.edu}}) is the best
means of contacting me. I will do my best to respond within 24 hours. If
I do not reply within 48 hours, do not take it personally, and
\emph{feel free to send a follow up email} in the very likely event that
I genuinely did not see your original message.

\hypertarget{office-hours}{%
\subsubsection{Office Hours}\label{office-hours}}

I am generally in my office Monday-Thursday during ``normal business
hours.'' You are always welcome to walk-in and chat about class,
college, careers, or anything at all. Please do try to use the official
office hours stated at the head of the syllabus if possible. If you need
to meet at a different time, I request that you send me an email or let
me know after class so I know when to expect you. If you want to go over
material from class, please have \emph{specific} questions you want help
with. I am not in the business of giving private lectures (particularly
if you missed class without a valid excuse).

Watch the excellent and accurate video explaining office hours (on
website syllabus page):

\hypertarget{enrollment}{%
\subsubsection{Enrollment}\label{enrollment}}

Students are responsible for verifying their enrollment in this class.
The last day to add or drop this class with no penalty is
\textbf{Tuesday, September 4}. Be aware of
\href{https://www.hood.edu/offices-services/registrars-office/academic-calendar}{important
dates}.

\hypertarget{honor-code}{%
\subsubsection{Honor Code}\label{honor-code}}

Hood College has an Academic Honor Code which requires all members of
this community to maintain the highest standards of academic honesty and
integrity. Cheating, plagiarism, lying, and stealing are all prohibited.
All violations of the Honor Code are taken seriously, will be reported
to appropriate authority, and may result in severe penalties, including
expulsion from the college. See
\href{http://hood.smartcatalogiq.com/en/2016-2017/Catalog/The-Spirit-of-Hood/The-Academic-Honor-Code-and-Code-of-Conduct}{here}
for more detailed information.

\hypertarget{van-halen-and-mms}{%
\subsubsection{Van Halen and M\&Ms}\label{van-halen-and-mms}}

When you have completed reading the syllabus, email me a picture of the
band Van Halen and a picture of a bowl of M\&Ms.~If you do this
\emph{before} the date of the first exam, you will get 3 bonus points on
the exam. Yes, this is real.

\hypertarget{accessibility-equity-and-accommodations}{%
\subsubsection{Accessibility, Equity, and
Accommodations}\label{accessibility-equity-and-accommodations}}

College courses can, and should, be challenging and bring you out of
your comfort zone in a safe and equitable environment. If, however, you
feel at any point in the semester that certain assignments or aspects of
the course will be disproportionately uncomfortable or burdensome for
you due to any factor beyond your control, please come see me or email
me. I am a very understanding person and am happy to work out a solution
together. I reserve the right to modify and reweight assignments at my
sole discretion for students that I belive would legitimately be at a
disadvantage, through no fault of their own, to complete them as
described.

If you are unable to afford required textbooks or other resources for
any reason, come see me and we can find a solution that works for you.

This course is intended to be accessible for all students, including
those with mental, physical, or cognitive disabilities, illness,
injuries, impairments, or any other condition that tends to negatively
affect one's equal access to education. If at any point in the term, you
find yourself not able to fully access the space, content, and
experience of this course, you are welcome to contact me to discuss your
specific needs. I also encourage you to contact the
\href{https://www.hood.edu/academics/josephine-steiner-center-academic-achievement-retention/accessibility-services}{Office
of Accessibility Services} (301-696-3421). If you have a diagnosis or
history of accommodations in high school or previous postsecondary
institutions, Accessibility Services can help you document your needs
and create an accommodation plan. By making a plan through Accessibility
Services, you can ensure appropriate accommodations without disclosing
your condition or diagnosis to course instructors.

\hypertarget{schedule}{%
\section{Schedule}\label{schedule}}

\textbf{You can find a full schedule} with resources for each class
meeting on the
\href{http://microF19.classes.ryansafner.com/schedule/}{course website
schedule page}.

\end{document}