---
title: "Practice Exam 1 Answer Key"
author: "Ryan Safner"
output:
  pdf_document:
    latex_engine: pdflatex #required for Fira Sans font to render properly 
header-includes:
  #- \usepackage{tikz, pgfplots, booktabs}
---

## Problem (30 points)

Show all work. You may *not* earn full credit if you only write the answer, even if correct.

11,. Suppose you can buy Drinks ($D$) and Wings ($W$) at a local bar.

a. **Suppose you have a budget of \$120/month to spend, the price of a Drink is \$3.00, and the price of a plate of Wings is \$4.00. Write a *graphable* equation for the budget constraint, and graph it on the first graph below. (3 points)**

---

$$\begin{aligned}
3D+4W&=120\\
4W&=120-3D\\
W&=30-0.75D\\
\end{aligned}$$

See original red line on first graph below.

---

b. **Suppose you earn utility according to the function:**

$$\begin{aligned}
u(D,W)&=DW\\
MU_D&=W\\
MU_W&=D\\
\end{aligned}$$

**Write an equation for your *marginal rate of substitution* between $D$ and $W$.  (3 points)**

---

$$\begin{aligned}
MRS_{D,W} &=\frac{MU_D}{MU_W}\\
MRS_{D,W} &=\frac{W}{D}\\
\end{aligned}$$

---

c. **Calculate the *optimum* quantities of $D$ and $W$ that maximizes your utility subject to your constraints. Plot this point on the first graph below, (call it point *A*), and sketch an indifference curve through that point. (10 points)**

---

$$\begin{aligned}
\frac{MU_D}{MU_W}&=\frac{p_D}{p_W}\\
\frac{W}{D}&=\frac{3}{4}\\
W&=\frac{3}{4}D\\
\end{aligned}$$

Plug this into the budget constraint:

$$\begin{aligned}
3D+4W&=120\\
3D+4(\frac{3}{4}D)&=120\\
3D+3D&=120\\
6D&=120\\
D^*&=20\\
\end{aligned}$$

Knowing $D^*=20$, we can find $W^*$: 

$$\begin{aligned}
W&=\frac{3}{4}D\\
W&=\frac{3}{4}(20)\\
W^*&=15\\
\end{aligned}$$

This point, $(D^*, W^*) = (20,15)$ is plotted as point $A$ on the first graph below.

---

d. **How much utility do you earn from the optimum bundle (point *A*)? (1 point)**

---

$$\begin{aligned}
u(D,W)&=DW\\
u(20,15)&=20*15\\
u(20,15)&=300\\
\end{aligned}$$

---

e. **Now suppose the price of Drinks ($D$) *rises* to \$4.00. Find the equation of the *new budget constraint* (in *graphable* form), and add it to the first graph below. (3 points)**

---

$$\begin{aligned}
4D+4W&=120\\
4W&=120-4D\\
W&=30-D\\
\end{aligned}$$

See new dotted red line on first graph below.

---

f. **Calculate the *new optimum* quantities of $x$ and $y$ under this new price. Plot this point on the first graph below, (call it point *B*), and sketch an indifference curve through that point. Hint: the formula for MRS has not changed. (6 points)**

---
$$\begin{aligned}
\frac{MU_D}{MU_W}&=\frac{p_D}{p_W}\\
\frac{W}{D}&=\frac{4}{4}\\
W&=D\\
\end{aligned}$$
Plug this into the (new!) budget constraint:

$$\begin{aligned}
4D+4W&=120\\
4D+4(D)&=120\\
4D+4D&=120\\
8D&=120\\
D^*_2&=15\\
\end{aligned}$$

Knowing $D^*_2=20$, we can find $W_2^*$: 

$$\begin{aligned}
W&=D\\
W^*_2&=15\\
\end{aligned}$$

This point, $(D^*_2, W^*_2) = (15,15)$ is plotted as point $B$ on the first graph below.

---

g. **How much utility do you earn from the new optimum bundle (point *B*)? (1 point)**

---

$$\begin{aligned}
u(D,W)&=DW\\
u(15,15)&=15*15\\
u(15,15)&=225\\
\end{aligned}$$

---

h. **Plot the optima from parts c and f on the *second* graph below, describing the relationship between the price of Drinks and the optimal consumption of Drinks. Connect the points. What did you just draw? (3 points)**

---

```{r, echo=F, warning=F, message=F}
library(tidyverse)
library(ggrepel)
points <-tribble(
  ~x, ~y, ~letter,
  20, 15, "A",
  15, 15, "B"
)
IC_1=function(x){300/x}
IC_2=function(x){225/x}

budget_1 = function(x){30-0.75*x}
budget_2 = function(x){30-x}

ggplot(tibble(x=c(0,10)), aes(x=x))+
  stat_function(fun=IC_1, geom="line", size=2, color = "green")+
  stat_function(fun=IC_2, geom="line", size=2, color = "blue")+
  stat_function(fun=budget_1, geom="line", size=2, color = "red")+
  stat_function(fun=budget_2, geom="line", size=2, color = "red", linetype="dashed")+
  geom_segment(x=0, xend=20,y=15,yend=15, linetype="dotted")+
  geom_segment(x=15, xend=15,y=15,yend=0, linetype="dotted")+
  geom_segment(x=20, xend=20,y=15,yend=0, linetype="dotted")+
  geom_point(data=points, aes(x=x,y=y), size=3)+
  geom_label_repel(data=points, aes(x=x,y=y, label=letter), size=5)+
    scale_x_continuous(breaks=seq(0,50,5),
                     limits=c(0,50),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,50,5),
                     limits=c(0,50),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "Drinks (D)",
       y = "Wings (W)")+
  theme_bw(base_size=16)
```

```{r, echo=F, warning=F, message=F}
library(tidyverse)
library(ggrepel)
points_2 <-tribble(
  ~x, ~y, ~letter,
  20, 3, "A",
  15, 4, "B"
)
demand=function(x){7-0.2*x}

ggplot(tibble(x=c(0,10)), aes(x=x))+
  stat_function(fun=demand, geom="line", size=2, color = "blue")+
  geom_label(aes(x=30, y=1),label = "Demand", color = "blue")+
  geom_segment(x=0, xend=20,y=3,yend=3, linetype="dotted")+
  geom_segment(x=20, xend=20,y=3,yend=0, linetype="dotted")+
  geom_segment(x=0, xend=15,y=4,yend=4, linetype="dotted")+
  geom_segment(x=15, xend=15,y=4,yend=0, linetype="dotted")+
  geom_point(data=points_2, aes(x=x,y=y), size=3)+
  geom_label_repel(data=points_2, aes(x=x,y=y, label=letter), size=5)+
    scale_x_continuous(breaks=seq(0,50,5),
                     limits=c(0,50),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,5,1),
                     limits=c(0,5),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "Drinks (D)",
       y = "Price per Drink (p)")+
  theme_bw(base_size=16)
```


