---
title: "Final Study Guide"
author: "Ryan Safner"
date: "12/9/2019"
header-includes:
  - \usepackage{color}
  - \usepackage{multirow}
  - \usepackage{multicol}
  - \usepackage{booktabs}
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
library(tidyverse)
library(mosaic)
```

# Market Equilibrium

## Demand and Supply

- **Demand function**: relates quantity demanded to price, e.g. 
$$q_D=12-2p$$
- **Inverse demand function**: relates price to quantity demanded, e.g. 
$$p=6-0.5q_D $$
  - Describes the ordinary graph of the demand curve: 

```{r, fig.align="center", fig.width=3, fig.asp=1}
demand=function(x){6-0.5*x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=demand, geom="line", size=2, color = "blue")+
  geom_label(aes(x=8,y=demand(8)), color = "blue", label="Demand")+
    scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)),
                     labels = function(x){paste("$", x, sep="")})+
  coord_cartesian(clip="off")+
  labs(x = "Quantity",
       y = "Price")+
  theme_light(base_size=14)
```

- *Choke price*: price where demand crosses the vertical axis $(q_D=0)$
- Can always obtain inverse demand function by solving for $p$ in the demand function

- **Supply function**: relates quantity supplied to price, e.g. 
$$q_S=0.5p-0.5$$
- **Inverse supply function**: relates price to quantity supplied, e.g. 
$$p=1+2q_S$$
  - Describes the ordinary graph of the supply curve:

```{r, fig.align="center", fig.width=3, fig.asp=1}
supply=function(x){1+2*x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=supply, geom="line", size=2, color = "red")+
  geom_label(aes(x=8,y=supply(8)), color = "red", label="Supply")+
    scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)),
                     labels = function(x){paste("$", x, sep="")})+
  coord_cartesian(clip="off")+
  labs(x = "Quantity",
       y = "Price")+
  theme_light(base_size=14)
```

- *Choke price*: price where demand crosses the vertical axis $(q_D=0)$
- Can always obtain inverse demand function by solving for $p$ in the demand function

## Equilibrium

```{r, fig.align="center", fig.width=3, fig.asp=1}
ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=supply, geom="line", size=2, color = "red")+
  geom_label(aes(x=8,y=supply(8)), color = "red", label="Supply")+
  stat_function(fun=demand, geom="line", size=2, color = "blue")+
  geom_label(aes(x=8,y=demand(8)), color = "blue", label="Demand")+
    scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)),
                     labels = function(x){paste("$", x, sep="")})+
  geom_segment(x=0,xend=2,y=5,yend=5,linetype="dashed",size=1)+
  geom_segment(x=2,xend=2,y=5,yend=0,linetype="dashed",size=1)+
  coord_cartesian(clip="off")+
  labs(x = "Quantity",
       y = "Price")+
  theme_light(base_size=14)
```

- Equilibrium exists at a unique price $p^*$ where $q^*=q_D=q_S$
- $p^*$ can always be found by setting original Demand function and Supply function

$$\begin{aligned}
		q_D&=q_S\\
		12-2p&=0.5p-0.5\\
		12&=2.5p-0.5\\
		12.5&=2.5p\\
		5&=p^*
		\end{aligned}$$

- Knowing $p^*$, can plug into either Demand function or Supply function to find $q^*$:
  
$$\begin{aligned}
		q_D&=12-2p\\
		q_D&=12-2(5)\\
		q^*&=2\\	
		\end{aligned}$$

## Disequilibrium: Surplus and Shortage

- **Shortage** (excess demand), a price below $p^*$, $q_D>q_S$
  - buyers will bid price upwards
- **Surplus** (excess supply), a price above $p^*$, $p_D<q_S$
  - sellers will lower asking prices

## Consumer and Producer Surplus

```{r, fig.align="center", fig.width=3, fig.asp=1}
cs<-tribble(
  ~x, ~y,
  0, 6,
  0, 5,
  2, 5
)

ps<-tribble(
  ~x, ~y,
  0, 1,
  0, 5,
  2, 5
)
ggplot(data.frame(x=c(0,10)), aes(x=x))+
  geom_polygon(data=cs, aes(x=x,y=y),fill="blue",alpha=0.4)+
  geom_polygon(data=ps, aes(x=x,y=y),fill="red",alpha=0.4)+
  stat_function(fun=supply, geom="line", size=2, color = "red")+
  geom_label(aes(x=8,y=supply(8)), color = "red", label="Supply")+
  stat_function(fun=demand, geom="line", size=2, color = "blue")+
  geom_label(aes(x=8,y=demand(8)), color = "blue", label="Demand")+
    scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)),
                     labels = function(x){paste("$", x, sep="")})+
  geom_segment(x=0,xend=2,y=5,yend=5,linetype="dashed",size=1)+
  geom_segment(x=2,xend=2,y=5,yend=0,linetype="dashed",size=1)+
  labs(x = "Quantity",
       y = "Price")+
  theme_light(base_size=14)
```

- Consumer Surplus $=$ Max WTP (Demand)$ - p^*$

- Producer Surplus $= p^* -$ Min WTA (Supply)$

- Area of Triangle $= \frac{1}{2}bh$

- Elasticity (in equilibrium) affects surplus:
  - *More* elastic:
    - less benefit from this particular exchange (have other options, etc)
    - less distance between Max WTP or Min WTA (choke price) and market price
    - less surplus
  - *Less* elastic:
    - more benefit from this particular exchange (have few options, etc)
    - greater distance between Max WTP or Min WTA (choke price) and market price
    - more surplus

## Efficiency of Markets

- Entrepreneurship, arbitrage, markets as a process

- Role of prices in coordinating information and incentives

- **Allocative efficiency**: allocate resources to highest-valued uses
  - maximum consumer and producer surplus

- **Pareto efficiency**: no improvements exist that would make at least one person better off without making another person worse off

- Markets are efficient when they
  1. Are competitive
  2. Can reach equilibrium
  3. Have no externalities

# Monopoly

## Features 

1. Firm's products may have few close substitutes

2. Barriers to entry, making entry costly

3. Firm is a "price-searcher": can set optimal price $p^*$ in addition to quantity $q^*$

## Marginal Revenue, Markup, and Price Elasticity

```{r, fig.align="center", fig.width=3, fig.asp=1}
demand_2=function(x){20-2*x}
MR=function(x){20-4*x}

#q_choke<-(a/b)
#MR_choke<-(a/(2*b))

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=demand_2, geom="line", size=1, color="blue")+
  geom_label(x=7,y=demand_2(7), label="Demand (slope=-b)", color="blue")+
  stat_function(fun=MR, geom="line", size=1, color="purple")+
  geom_label(x=4.5,y=MR(4.5), label="MR(q) (slope=-2b)", color="purple")+
    scale_x_continuous(breaks = c(0,5,10),
                       labels = c(0,expression(a/2*b),expression(a/b)),
                       expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0,20),
                     labels=c(0,expression(a)),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "Quantity (q)",
       y = "Price (p)")+
  theme_classic(base_size=14)
```

- Inverse demand: $p=a-bQ$ $\implies$ Marginal revenue: $MR(q)=a-2bq$

| Price Elasticity | $MR(q)$ | $R(q)$ |
|------------|---------|--------|
| $\vert \epsilon \vert >1$  Elastic | $+$ | Increasing |
| $\vert \epsilon \vert =1$ Unit | $0$ | Maximized |
| $\vert \epsilon \vert <1$ Inelastic | $-$ | Decreasing |

- Size of markup depends on **price elasticity of demand**
  - $\downarrow$ price elasticity: $\uparrow$ markup

- **Lerner Index** measures market power as % of firm's price that is markup above (marginal) cost
$$L=\frac{p-MC(q)}{p} = -\frac{1}{\epsilon}$$
  - In perfect competition, $L=0$ (as $p=MC)$
  - As $L \rightarrow 1$, more market power

## Profit-Maximization Problem Solution

1. Produce the optimal amount of output $q^*$ where $MR(q)=MC(q)$
2. Raise price to maximum consumers are WTP: $p^*=Demand(q^*)$
3. Calculate profit with average cost: $\pi=[p-AC(q)]q$
4. Shut down in the *short run* if $p<AVC(q)$
    - Minimum of AVC curve where $MC(q)=AVC(q)$
5. Exit in the *long run* if $p<AC(q)$
    - Minimum of AC curve where $MC(q)=AC(q)$

## Consequences of Market Power

\begin{center}
  \includegraphics[height=2in]{pc3.png}
\end{center}

- In a *competitive* market in long run equilibrium:
  - **Economic profit** is driven to $0
  - **Allocatively efficient**: $p=MC(q)$ (goods produced until $MB=MC)$
  - **Productively efficient**: $p=AC(q)_{min}$, otherwise firms would enter/exit
  - Consumer surplus and producer surplus is maximized

\begin{center}
  \includegraphics[height=2in]{pc4.png}
\end{center}

- If that same market were monopolized:
  - Monopolist sets $MR(q)=MC(q)$, raises price to Max WTP (Demand)
  - Restricts output and raises price, compared to competitive market
  - Earns monopoly profits $(p>AC)$
  - Loss of consumer surplus
  - **Deadweight loss** of surplus destroyed from lost gains from trade

- Rent-seeking
  - "prize" of monopoly is monopoly profits
  - firm(s) willing to invest resources to compete for the privilege to be a monopoly (e.g. lobbying for barriers to entry, preventing competition, etc)

## Sources of Market Power

1. Control over a key resource

2. Barriers to entry
  - ex: occupational licensing, intellectual property rights, anticompetitive regulation, etc.

3. Economies of scale/**natural monopoly**

\begin{center}
  \includegraphics[height=2in]{naturalmon.png}
\end{center}

- One firm with greater economies of scale can produce more at a lower cost than competition
  - Often regulated by government - force the monopolist to act closer to a competitive outcome $(p=MC)$


# Pricing Strategies

- Goal of price-discrimination is to charge different prices to different customers to convert consumer surplus into profit for firm

- To engage in price discrimination, two conditions:

1. Firm must have market power
2. Firm must be able to prevent arbitrage/resale

- 1^st^-degree price discrimination: firm charges each customer their max WTP

- 3^rd^-degree price discrimination: firm segments market into multiple groups based on demand/elasticity differences
  - charge higher price to less-elastic group
  - charge lower price to more-elastic group
  - must be able to separate customers into groups by identifiable characteristics before sale

- 2^nd^-degree price discrimination: firm can't identify customer type beforehand, offers different options
  - tying: lower price on "base" good, raise price on refills
  - bundling: combine multiple goods into a package and prevent sale of individual components of bundle

# Monopolistic Competition

## Features 

- Firms have some market power

1. Firms selling imperfect substitutes

2. No Barriers to entry

3. Firm is a "price-searcher"

- In the short run, modeled like a monopoly

\begin{center}
  \includegraphics[height=2in]{naturalmon.png}
\end{center}

- In the long run, no barriers to entry $\implies$ competitive entry pushes $\pi$ to 0 
  - demand for each firm's product decreases & becomes more elastic until $p=AC$ for each firm


\begin{table}
  \begin{tabular}{cc}
  \includegraphics[height=2in]{mcpc.png} &
  \includegraphics[height=2in]{mcpc2.png}\\
  \end{tabular}
\end{table}

- Compare to perfect competition (left)
  - Lower output and higher price, less consumer surplus, some deadweight loss
  - Worse than perfect competition, but better than monopoly

## Oligopoly

- Industry with few sellers

- Firms are strategic and interdependent

- Prisoner's Dilemma: game where each player faces an incentive not to cooperate, but all players are better off if they all cooperate

	\begin{figure}[h!]
\centering 
\begin{tabular}{lrcc|cc|}
& & \multicolumn{4}{c} {\textbf{\textcolor{blue}{Player 2}}} \\
& & \multicolumn{2}{c}{Cooperate} & \multicolumn{2}{c}{Defect} \\ \cline{3-6}
\parbox[t]{2mm}{\multirow{4}{*}{\rotatebox[origin=c]{90}{\textbf{\textcolor{red}{Player 1}}}}}  & 
\multicolumn{1}{c|}{\multirow{2}{*}{Cooperate}} & \textcolor{red}{10} & & \textcolor{red}{0} & \\ 
&  &\multicolumn{1}{|c}{} & \textcolor{blue}{10} & & \textcolor{blue}{25}\\ \cline{3-6}
&
\multicolumn{1}{c|}{\multirow{2}{*}{Defect}} & \textcolor{red}{25} &  & \textcolor{red}{5} & \\ 
& & \multicolumn{1}{|c}{} & \textcolor{blue}{0} &  & \textcolor{blue}{5} \\ \cline{3-6}
\end{tabular}
\caption{Prisoner's Dilemma example}  
\end{figure}

- **Nash equilibrium**: outcome where each player has no incentive to switch strategies
  - In the example above, it is (Defect, Defect)

- **Cartel**: firms colluding to raise prices together and split monopoly profits
  - Not a Nash equilibrium! Each player has an incentive to break the agreement and Defect

# Comparing Industries

| Industry | Firms | Entry | Price (LR Eq.) | Output | Profits (LR) | Cons. Surplus | DWL |
|----------|-------|-------|--------|--------|--------------|---------------|-----|
| Perfect competition | Very many | Free | Lowest $(MC)$ | Highest | 0 | Highest | None |
| Monopolistic competition | Many | Free | Higher $(p>MC)$ | Lower | 0 | Lower | Some |
| Oligopoly (non-cooperative) | Few | Barriers? | Higher | Lower | Some | Lower | Some |
| Monopoly<sup>1</sup> (or cartel) | 1 | Barriers | Highest | Lowest | Highest | Loweset | Largest |

# Contestable Markets

- Markets are **contestable** if:
  1. There are no barriers to entry or exit
  2. Firms have similar technologies (i.e. similar cost structure)
  3. There are no sunk costs

- Threat of entry $\implies$ Nash equilibrium is the competitive outcome, $p=MC$ with just 1 firm!

