---
title: "1.5 Preferences II: MRS and Marginal Utility - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Overview

We continue with our model of how individuals choose. Our focus for this class is to wrap up what we mean by "maximizing utility" as a person's *objective.* We will do some practice problems on understanding indifference curves and utility functions. Then we will finally be ready to put our tools together next class to solve the consumer's problem.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/05-slides.html)

# Practice Problems

Today you will be working on [practice problems](/practice/05-practice.html). Answers will be posted later.

# Math Appendix

## Graphing Indifference Curves

I will not ask you to formally graph indifference curves (just roughly sketch them where appropriate). If you wanted to graph them, and express them in a graphable (slope-intercept form) equation, simply solve for the good on the vertical axis. 

**Example**: Suppose we have a typical^[Cobb-Douglas!] indifference curve for apples $(a)$ and bananas $(b)$: 
$$u(a,b)=ab$$

```{r, fig.retina=3, fig.height=5.5}
library("plotly")
x_vec = seq(0,10,1)
y_vec = seq(0,10,1)
x_matrix = matrix(c(x_vec), nrow = 10, ncol = 1)
y_matrix = matrix(c(y_vec), nrow = 1, ncol = 10)
z_matrix = matrix(c(x_matrix %*% y_matrix),nrow=10,ncol=10)

plot_ly(x = x_vec,
           y = y_vec,
           z = z_matrix) %>%
  add_surface() %>%
  layout(scene = list(xaxis = list(title="a"),
                      yaxis = list(title="b"),
                      zaxis = list(title="Utility")))
```

Each indifference curve (or contour) is one level of utility (all points on the curve give a specific level of utility). So, set that level of utility equal to some constant, $k$. 

$$ab=k$$

Then, if we are putting $b$ on the vertical axis, we simply solve this for $b$:

$$\begin{align*}
ab&=k\\
b&=\frac{k}{a}\\
\end{align*}$$

This is the general equation for all indifference curves of this utility function: each utility level (value of $k)$ can be graphed as an indifference curve with that equation. Thus, for example, for a utility level of $10$, the equation for that indifference curve is
$$b=\frac{10}{a}$$.

```{r, fig.retina=3, fig.height=5.5}
library("mosaic")
update_geom_defaults("label", list(family = "Fira Sans Condensed"))

IC_50=function(x){50/x}
IC_25=function(x){25/x}
IC_10=function(x){10/x}

colors<-c("b=50/a" = "#20A387FF", "b=25/a" = "#39568CFF", "b=10/a" = "#481567FF")

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=IC_50, geom="line", size=2, aes(color = "b=50/a"))+
  stat_function(fun=IC_25, geom="line", size=2, aes(color = "b=25/a"))+
  stat_function(fun=IC_10, geom="line", size=2, aes(color = "b=10/a"))+
  geom_label(aes(x=9,y=IC_50(9)), color = "#20A387FF", label="u=50", size = 6)+
  geom_label(aes(x=9,y=IC_25(9)), color = "#39568CFF", label="u=25", size = 6)+
  geom_label(aes(x=9,y=IC_10(9)), color = "#481567FF", label="u=10", size = 6)+
    scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_colour_manual("I.C. equations", values = colors)+
  labs(x = "a",
       y = "b",
       title = "Indifference curves for u(a,b)=ab")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

## Cobb-Douglas Functions

Cobb-Douglas functions (both for utility and production) are one of the most common functional forms in economics. Despite their somewhat frightening presence on the surface, they have very neat mathematical properties, and have been empirically useful as well. A Cobb-Douglas function (for utility) looks like: 

$$u(x, y)=x^a y^b$$
		
### MRS and Positive Monotonic Transformations

The marginal rate of substitution, using the above equation is: 
$$\begin{align*}
	MRS_{x, y}	&=\frac{MU_{x}}{MU_{y}} && \text{The definition of MRS}\\
	&= \cfrac{\big(\frac{\partial }{\partial x}\big)}{\big(\frac{\partial}{\partial y}\big)} && \text{Using the definition of marginal utility as partial derivative}\\
	&= \frac{a x^{a-1}y^b}{b x^a y^{b-1}} && \text{Differentiating }u \text{ w.r.t }x \text{ on top and }y \text{ on bottom}\\
	&=\frac{a}{b} x^{(a-1)-a}y^{b-(b-1)} && \text{Using exponent rules for division}\\
	&= \frac{a}{b} x^{-1}y^{1} && \text{Simplifying exponents}\\
	&=\frac{a}{b} \frac{y}{x} && \text{Using exponent rules for negative exponents}\\ \end{align*}$$

Using logarithms, we can take a positive monotonic transformation of the original utility function $u$:

$$v(x, y)=a \: ln \: x + b \: ln \: y$$

The MRS using the logarithmic form is:

$$\begin{align*}
	MRS_{x, y}	&=\frac{MU_{x}}{MU_{y}} && \text{The definition of MRS}\\
	&= \cfrac{\big(\frac{\partial }{\partial x}\big)}{\big(\frac{\partial}{\partial y}\big)} && \text{Using the definition of marginal utility as partial derivative}\\
	&= \cfrac{\big(\frac{a}{x}\big)}{\big(\frac{b}{y}\big)} && \text{Differentiating }u \text{ w.r.t }x \text{ on top and }y \text{ on bottom}\\
	&=\frac{a}{b} \frac{y}{x} && \text{Multiplying the righthand side by the reciprocal of the denominator,} \frac{y}{b}\\ \end{align*}$$
	
Which of course, is the same, because we know any positive monotonic transformation of a utility function preserves the same preferences! 

### EXAMPLE

Find the marginal rate of substitution for the utility function: 

$$u(x, y) = \sqrt{x, y}$$

First, recognize that square roots are equivalent to saying $x^{0.5} y^{0.5}$. The marginal rate of substitution would be: 

$$\begin{align*}
		MRS_{x, y}	&=\frac{MU_{x}}{MU_{y}} && \text{The definition of MRS}\\
	&= \cfrac{\big(\frac{\partial }{\partial x}\big)}{\big(\frac{\partial}{\partial y}\big)} && \text{Using the definition of marginal utility as partial derivative}\\
	&= \frac{0.5 x^{-0.5}y^{0.5}}{0.5x^{0.5}y^{-0.5}} && \text{Differentiating }u \text{ w.r.t }x \text{ on top and }y \text{ on bottom}\\
	&=\frac{0.5}{0.5} x^{0.5-0.5}y^{0.5-(-0.5)} && \text{Using exponent rules for division}\\
	&=  x^{-1}y^{1} && \text{Simplifying exponents, and cancelling} \frac{0.5}{0.5}\\
	&=\frac{y}{x} && \text{Using exponent rules for negative exponents}\\ \end{align*}$$

Note we could find this equivalently again by taking logs of the original utility function. 

$$v(x,y)=0.5 \: ln \: x+0.5\: ln\: y$$
 	
$$\begin{align*}
		MRS_{x, y}	&=\frac{MU_{x}}{MU_{y}} && \text{The definition of MRS}\\
	&= \cfrac{\big(\frac{\partial }{\partial x}\big)}{\big(\frac{\partial}{\partial y}\big)} && \text{Using the definition of marginal utility as partial derivative}\\
	&= \cfrac{\big(\frac{0.5}{x}\big)}{\big(\frac{0.5}{y}\big)} && \text{Differentiating }u \text{ w.r.t }x \text{ on top and }y \text{ on bottom}\\
	&=\frac{y}{x} && \text{Multiplying the righthand side by the reciprocal of the denominator,} \frac{y}{0.5}\\ \end{align*}$$

Which again, gives us the same marginal rate of substitution.

