---
title: "1.8 Income and Substitution Effects - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Overview

Today we cover the third effect: the (own) price effect, aka the law of demand: how changes in a good's (own) price affect quantity demanded for that good. We introduce the idea of demand functions, inverse demand functions, and a demand curve. I will also show you examples of how we derive demand functions and curves.

The other major concept today is breaking apart the law of demand into two effects: the (real) income effect, and the substitution effect. We considering them conceptually and graphically. These can be a difficult concept for students to grasp.

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/08-slides.html)

# Interactive Examples

- [Visualizing the Consumer's Problem](https://ryansafner.shinyapps.io/Consumer/)
- [Visualizing Changes in the Consumer's Problem](https://ryansafner.shinyapps.io/Consumer_changes/)
- [Visualizing Demand Shifters](https://ryansafner.shinyapps.io/Demand/)

These are examples that I wrote in R/Shiny in the past, to visualize what it is we are looking at. As I find more time, I will update these and integrate them into our slides, but until then, I will just post links. 

The first is a visual example of the static (one-time) Consumer's problem. You can adjust things about the (Cobb-Douglas) utility function, income, and prices, and see how it would affect the graph and the optimum.

The second is a visual example of dynamic changes in the Consumer's problem. The consumer starts at a pre-defined optimum. You can adjust any of the budget constraint parameters $(m, p_x, p_y)$, and see how it would affect a new optimum, which you can *compare* the the original optimum.

The third is a visual example of a hypothetical demand function for beer, with preference intensity, income, price of nachos (a complement), and price of wine (a substitute) as inputs to the function. Based on some pre-defined functions (in the background), you can change any of the inputs and see how it would affect the (inverse) demand function, expressed as a simple demand curve. 

I hope some time to clean these up and make a few more, such as "Visualizing the Income and Substitution Effects" and "Estimating a Demand Function from Real World Data."

# Practice Problems

There are no official practice problems for today, but you should try to draw the full (real) income and substitution effects graph for a price change (for a normal good) on your own!

# Assignments: Problem Set 1 is DUE and Problem Set 2

[Problem set 1](/assignment/01-problem-set) is DUE!

Problem set 2 (on classes 1.7-1.9) will be posted shortly, and is due by Thursday September 26 (for section 1 - TuTh) or Monday September 30 (for section 2 - MW).

# Math Appendix

## Example Demand Functions

The examples below show how we can take a utility function and derive a demand function.

### Perfect Substitutes

Recall that perfect substitutes have indifference curves that are straight lines (constant MRS), and have a utility function of:
$$u(x,y)=w_1x+w_2y$$

Since both the budget constraint and the indifference curves are straight lines, we will not have a point of **tangency** between them, but a point(s) where the lines **intersect**.

```{r, fig.retina=3, fig.height=5.5}
library("mosaic")
library("ggrepel")
update_geom_defaults("label", list(family = "Fira Sans Condensed"))

IC_s1=function(x){6-x}
IC_s2=function(x){8-x}
IC_s3=function(x){10-x}
BC=function(x){4-0.5*x}

point_sub<-tribble(
  ~letter, ~x, ~y,
  "A", 8, 0
)
#colors<-c("b=50/a" = "#20A387FF", "b=25/a" = "#39568CFF", "b=10/a" = "#39568CFF")

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=IC_s1, geom="line", size=2, color = "#481567FF")+
  stat_function(fun=IC_s2, geom="line", size=2, color = "#39568CFF")+
  stat_function(fun=IC_s3, geom="line", size=2, color = "#20A387FF")+
  stat_function(fun=BC, geom="line", size=2, color = "red")+
  geom_label(aes(x=2,y=IC_s1(2)), color = "#481567FF", label="u=6", size = 6)+
  geom_label(aes(x=2,y=IC_s2(2)), color = "#39568CFF", label="u=8", size = 6)+
  geom_label(aes(x=2,y=IC_s3(2)), color = "#20A387FF", label="u=10", size = 6)+
  geom_label(aes(x=1,y=BC(1)), color = "red", label="BC", size = 6)+
  geom_point(data = point_sub,
             mapping=aes(x=x, y=y),
             size = 4)+
  geom_text_repel(data = point_sub,
             mapping=aes(x=x, y=y,
             label = letter),
             size = 6)+
    scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  coord_cartesian(clip="off")+
  labs(x = "a",
       y = "b",
       title = "Optimal Consumption for Substitutes",
      caption = expression(paste(u(x,y)==x+y, ", ", m, " = $20.00, ", p[x]," = $2.50, ", p[y]," = $5.00", sep=" ")))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

There are three possible cases, 

1. if $p_x>p_y$, the slope of the budget constraint is flatter than the slope of the indifference "curves." The optimal bundle is spending *all* income on good $x$.
2. if $p_x<p_y$, the slope of the budget constraint is steeper than the slope of the indifference "curves." The optimal bundle is spending *all* income on good $y$.
3. if $p_x=p_y$, the budget constraint and indifference "curves" are *the same line*, so *any point* on the lines is optimal!

The first two cases are known as "**corner solutions**," where a consumer chooses all of one good, and none of another. Note perfect substitutes are *not convex*, and violate the assumption that "averages are preferred to extremes."

We can therefore write the **demand function for good x** (and similarly for y):

$$q_d^x = \begin{cases} \frac{m}{p_x} & \text{When } p_x < p_y\\
\text{any number between 0 and } \frac{m}{p_x} & \text{When } p_x=p_y\\
0 & \text{When } p_x > p_y\\ \end{cases}$$

### Perfect Complements

Recall that perfect complements have indifference curves that are right angles, and have a utility function of:
$$u(x,y)=\min \{w_1x,w_2y\}$$

The optimal choice must always be at the **corner** of the indifference "curves."

```{r, fig.retina=3, fig.height=5.5}
library("mosaic")
library("ggrepel")
update_geom_defaults("label", list(family = "Fira Sans Condensed"))

BC_c=function(x){6-x}

point_comp<-tribble(
  ~letter, ~x, ~y,
  "A", 3, 3
)

ggplot(data.frame(x=c(0,5)), aes(x=x))+
  stat_function(fun=BC_c, geom="line", size=2, color = "red")+
  geom_label(aes(x=1,y=BC_c(1)), color = "red", label="BC", size = 6)+
  geom_segment(aes(x=2, xend=2, y=10, yend=2), size=1, color="#453781FF")+
  geom_segment(aes(x=2, xend=10, y=2, yend=2), size=1, color="#453781FF")+
  geom_segment(aes(x=3, xend=3, y=10, yend=3), size=1, color="#20A387FF")+
  geom_segment(aes(x=3, xend=10, y=3, yend=3), size=1, color="#20A387FF")+
  geom_label(aes(x=9, y=2), size=6, color="#453781FF", label="u=2")+
  geom_label(aes(x=9, y=3), size=6, color="#20A387FF", label="u=3")+
  geom_point(data = point_comp,
             mapping=aes(x=x, y=y),
             size = 4)+
  geom_text_repel(data = point_comp,
             mapping=aes(x=x, y=y,
             label = letter),
             size = 6)+
  scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "x",
       y = "y",
       title = "Optimal Consumption for Complements",
      caption = expression(paste(u(x,y)==min(x,y), ", ", m, " = $20.00, ", p[x]," = $2.50, ", p[u]," = $2.50", sep=" ")))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

The consumer is purchasing equal amounts of $x$ and $y$ in this example, no matter what the prices are of either, so $x=y$. We must satisfy the budget constraint, that $p_xx+p_yx=m$ (recall $y=x$). Solving this for $x$ gives us the **demand function for good x**:

$$q_D^x=q_D^y=\frac{m}{p_x+p_y}$$

This should be intuitive: the consumer must always purchase equal amounts of $x$ and $y$, so it's as if the consumer was spending all of their income on a single good $(x+y)$ that has a price of $(p_x+p_y)$.

## Cobb-Douglas

Now we come to a more realistic demand function. If a consumer has a Cobb-Douglas utility function^[Note: Natural logs [are easier to work with](https://microf19.classes.ryansafner.com/class/05-class/#cobb-douglas-functions), this function is equivalent to $u(x,y)=x^ay^b$.]
$$u(x,y)= a \ln x+ b \ln y$$

Then we are solving the following constrained optimization problem:

$$\max_{x,y} a \ln x + b \ln y$$
$$\text{subject to } p_xx+p_yy=m$$

I will solve this using the Lagrangian method.^[Note you could solve this by substitution. Use the definition of the optimum, that $\frac{MU_x}{MU_y}=\frac{p_x}{p_y}$, plug in, and solving carefully for $x$.]

$$\mathbb{L}=a \: ln \: x + b \: ln \: y - \lambda(p_x x + p_y y - m)$$

The First Order Conditions are:

$$\begin{align*}
\frac{\partial \mathbb{L}}{\partial x} = \frac{a}{x}-\lambda p_x &= 0\\
\frac{\partial \mathbb{L}}{\partial y} = \frac{b}{y}-\lambda p_y &= 0\\
\frac{\partial \mathbb{L}}{\partial \lambda} = p_xx+p_yy-m &= 0\\ \end{align*}$$

Rearranging the first two FOCs:

$$\begin{align*}
a&=\lambda p_x x\\
b&=\lambda p_y y\\
\end{align*}$$

Adding them together:

$$\begin{align*}
a+b&=\lambda(p_xx+p_yy)\\
a+b&=\lambda m\\ \end{align*}$$

Solving for $\lambda$: 

$$\lambda=\frac{a+b}{m}$$

Substitute this back into each of the first two FOCs and solve for $x$ and $y$, respectively, to get:

$$\begin{align*}
x&=\frac{a}{a+b}\frac{m}{p_x}\\
y&=\frac{b}{a+b} \frac{m}{p_y}\\
\end{align*}$$

These are the **demand functions** for goods $x$ and $y$. One of the **convenient properties** of Cobb-Douglas preferences is that a consumer *spends a constant fraction of income* $(\frac{a}{a+b})$ *on good $x$*, and *the complement fraction* $(\frac{b}{a+b})$ *on good $y$*!^This becomes especially simple when the exponents $a+b=1$, as in:

$$u(x,y)=x^ay^{(1-a)} \quad \quad \text{where 0 < a < 1}$$

So $a$ and $1-a$ are proportions of $m$!]

For a constant value of $p_x$, this is a linear function of $m$. So doubling, tripling, etc. $m$ will double, triple, etc. quantity demanded for $x$. So the income expansion path ([see next class](/class/08-class)) is a straight line through the origin with a slope $\frac{a}{p_x}$.

### EXAMPLE

Suppose a consumer has a utility function

$$u(x,y)=x^{0.5}y^{0.5}$$

and faces constraints of $m=50$, $p_a=10$, $p_b=5$.

$$\max_{x,y} x^{0.5}y^{0.5}$$
$$\text{subject to } 10x+5y=50$$

Write the Lagrangian (and taking logs of the objective function, for simplicity):

$$\mathbb{L} = 0.5 \: ln \: x+0.5 \: ln \: y-\lambda(10x+5y-50)$$

FOCs are:

$$\begin{align*}
\frac{\partial \mathbb{L}}{\partial x} = \frac{0.5}{x}-10\lambda &= 0 \\
\frac{\partial \mathbb{L}}{\partial y} = \frac{0.5}{y}-5\lambda &= 0 \\
\frac{\partial \mathbb{L}}{\partial \lambda} = 10x+5y-50 &= 0\\
\end{align*}$$

Rearrange the first two FOCs:

$$\begin{align*}
0.5&=10 \lambda x\\
0.5&=5 \lambda y\\
\end{align*}$$

Add then together, and solve for $\lambda$:

$$\begin{align*}
1&=\lambda(10x+5y)\\
1&=\lambda(50)\\
\frac{1}{50}&=\lambda\\
\end{align*}$$

Now substitute lambda into each of the first two FOCs, solving for $x$ and $y$, respectively: 

$$\begin{align*}
\frac{0.5}{x}&=10(\frac{1}{50})\\
10x&=0.5(50)\\
x&=2.5\\
\end{align*}$$

$$\begin{align*}
\frac{0.5}{y}&=5(\frac{1}{50})\\
10x&=0.5(50)\\
y&=5\\
\end{align*}$$

This consumer buys 2.5 units of $x$ at $10/each ($25) and 5 units of $y$ at $5/each ($25), spending all $50. 
	
Recall the utility function is

$$u(x, y)=\sqrt{xy}=x^{0.5}y^{0.5}$$

Note the Cobb-Douglas exponents on $x$ and $y$ are equal and sum to 1: $(a$ and $1-a=0.5)$. So the consumer spends $a=50\%$ of her income on $x$, and $1-a=50\%$ on $y$. 

To get the **demand functions** for $x$ and $y$, go back to the FOCs, and keep all budget constraint terms as variables $(p_x, p_y, m)$.

$$\begin{align*}
\frac{\partial \mathbb{L}}{\partial x} = \frac{0.5}{x}-p_x\lambda &= 0 \\
\frac{\partial \mathbb{L}}{\partial y} = \frac{0.5}{y}-p_y\lambda &= 0 \\
\frac{\partial \mathbb{L}}{\partial \lambda} = p_xx+p_yy-m &= 0\\
\end{align*}$$

Rearrange the first two FOCs:

$$\begin{align*}
0.5&=p_x \lambda a\\
0.5&=p_y \lambda y\\
\end{align*}$$

Add then together, and solve for $\lambda$:

$$\begin{align*}
1&=\lambda(p_xx+p_yy)\\
1&=\lambda(m)\\
\frac{1}{m}&=\lambda\\
\end{align*}$$

Substitute this back into each of the first two FOCs and solve for $x$, and $y$, respectively, to get:

$$\begin{align*}
\frac{0.5}{x}-p_x\lambda&=0\\
\frac{0.5}{x}-p_x \left(\frac{1}{m}\right)&=0\\
\frac{0.5}{x}&=\frac{p_x}{m}\\
p_xx&=0.5m\\
x&=0.5\frac{m}{p_x}\\
\end{align*}$$

$$\begin{align*}
\frac{0.5}{y}-p_y\lambda&=0\\
\frac{0.5}{y}-p_y \left(\frac{1}{m}\right)&=0\\
\frac{0.5}{y}&=\frac{p_y}{m}\\
p_yy&=0.5m\\
y&=0.5\frac{m}{p_y}\\
\end{align*}$$

These are the **demand functions** for $x$ and for $y$.^[Note you can just take the rule learned above, that $x=a\frac{m}{p_x}$ and $y=b\frac{m}{p_y}$ and plug in $a$ and $b$.]


Note if we want to *graph* this, we need to find the **inverse demand function** by solving for $p_x$:

$$\begin{align*}
x&=0.5\frac{m}{p_x} \\
p_xx&=0.5m\\
p_x&=0.5\frac{m}{x} \\
\end{align*}$$

Now we can graph a demand function for a given amount of income. As income changes, the demand curve shifts. Here are two examples, one where $m=20$ and another where $m=40$: 

```{r, fig.retina=3, fig.height=5.5}
library("mosaic")
library("ggrepel")

cd1=function(x){0.5*20/x}
cd2=function(x){0.5*40/x}

point_sub<-tribble(
  ~letter, ~x, ~y,
  "A", 8, 0
)

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=cd1, geom="line", size=2, color = "blue")+
  stat_function(fun=cd2, geom="line", size=2, color = "blue", linetype="dashed")+
  geom_label(aes(x=3,y=cd1(3)), color = "blue", label="Demand: m=20", size = 6)+
  geom_label(aes(x=6,y=cd2(6)), color = "blue", label="Demand: m=40", size = 6)+
    scale_x_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,10,1),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)),
                     labels = function(x){paste("$", x, sep="")})+
  coord_cartesian(clip="off")+
  labs(x = "Quantity of x",
       y = "Price of x",
       title = "Demand Curve",
      caption = expression(p_x==0.5*m/x))+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```
