---
title: "2.4 Cost and Revenue Functions - Class Notes"
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-syllabus-no-bib.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Overview

Today we cover costs and revenues before we put them together next class to solve the firm's *profit maximization problem*. 

# Slides

- [<i class="fas fa-external-link-square-alt"></i> Lecture Slides](/slides/12-slides.html)

# Appendix

## The Relationship Between Returns to Scale and Costs

There is a direct relationship between a technology's returns to scale^[Increasing, decreasing, or constant] and its cost structure: the rate at which its total costs increase^[At a decreasing rate, at an increasing rate, or at a constant rate, respectively] and its marginal costs change^[Decreasing, increasing, or constant, respectively]. This is easiest to see for a single input, such as our assumptions of the short run (where firms can change $l$ but not $\bar{k})$:

$$q=f(\bar{k},l)$$

### Constant Returns to Scale: 

```{r, fig.retina=3, fig.height=5.5}
library(mosaic)
crs_prod=function(x){x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=crs_prod, geom="line", size=2, color="red")+
  geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
  # first dotted lines
  geom_segment(x=1, xend=1, y=0, yend=1, linetype="dotted",size=1)+
  geom_segment(x=0, xend=1, y=1, yend=1, linetype="dotted",size=1)+
  # second dotted lines
  geom_segment(x=2, xend=2, y=0, yend=2, linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=2, yend=2, linetype="dotted",size=1)+
    scale_x_continuous(breaks=c(0,1,2),
                       labels=c(0,"l","2l"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0,1,2),
                     labels=c(0,"f(l)", "f(2l)"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Input, Labor (l)",
       y = "Output (q)",
       title = "Constant Returns to Scale")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

```{r, fig.retina=3, fig.height=5.5}
crs_cost=function(x){x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=crs_cost, geom="line", size=2, color="red")+
  #geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y = "Total Costs C(q)",
       title = "Total Costs with Constant RTS")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

```{r, fig.retina=3, fig.height=5.5}
crs_mc=function(x){2}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=crs_mc, geom="line", size=2, color="red")+
  geom_label(aes(x=2.5,y=2), color = "red", label="MC(q)=AC(q)", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs per Unit",
       title = "Per Unit Costs with Constant RTS")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

### Decreasing Returns to Scale

```{r, fig.retina=3, fig.height=5.5}
library(mosaic)
drs_prod=function(x){sqrt(x)}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=drs_prod, geom="line", size=2, color="red")+
  geom_abline(intercept=0, slope=1, size=1, linetype="dashed")+
  geom_label(aes(x=2.5,y=drs_prod(2.5)), color = "red", label="Total Product", size = 5)+
  geom_label(aes(x=3,y=3), label=expression(45*degree), size = 5)+
  # first dotted lines
  geom_segment(x=1, xend=1, y=0, yend=1, linetype="dotted",size=1)+
  geom_segment(x=0, xend=1, y=1, yend=1, linetype="dotted",size=1)+
  # second dotted lines
  geom_segment(x=2, xend=2, y=0, yend=2, linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=2, yend=2, linetype="dotted",size=1)+
  # function dotted lines
  geom_segment(x=2, xend=2, y=0, yend=drs_prod(2), linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=drs_prod(2), yend=drs_prod(2), linetype="dotted",size=1)+
    scale_x_continuous(breaks=c(0,1,2),
                       labels=c(0,"l","2l"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0,1,drs_prod(2),2),
                     labels=c(0,"f(l)", "2f(l)", "f(2l)"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Input, Labor (l)",
       y = "Output (q)",
       title = "Decreasing Returns to Scale")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

```{r, fig.retina=3, fig.height=5.5}
drs_cost=function(x){x^{1.5}}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=drs_cost, geom="line", size=2, color="red")+
  #geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y = "Total Costs C(q)",
       title = "Total Costs with Decreasing RTS")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

```{r, fig.retina=3, fig.height=5.5}
drs_mc=function(x){x^{1.5}}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=drs_mc, geom="line", size=2, color="red")+
  geom_label(aes(x=2,y=drs_mc(2)), color = "red", label="MC(q)", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs per Unit",
       title = "Per Unit Costs with RTS")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

### Increasing Returns to Scale

```{r, fig.retina=3, fig.height=5.5}
library(mosaic)
irs_prod=function(x){x^{1.5}}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=irs_prod, geom="line", size=2, color="red")+
  geom_abline(intercept=0, slope=1, size=1, linetype="dashed")+
  geom_label(aes(x=2.5,y=irs_prod(2.5)), color = "red", label="Total Product", size = 5)+
  geom_label(aes(x=3,y=3), label=expression(45*degree), size = 5)+
  # first dotted lines
  geom_segment(x=1, xend=1, y=0, yend=1, linetype="dotted",size=1)+
  geom_segment(x=0, xend=1, y=1, yend=1, linetype="dotted",size=1)+
  # second dotted lines
  geom_segment(x=2, xend=2, y=0, yend=2, linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=2, yend=2, linetype="dotted",size=1)+
  # function dotted lines
  geom_segment(x=2, xend=2, y=0, yend=irs_prod(2), linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=irs_prod(2), yend=irs_prod(2), linetype="dotted",size=1)+
    scale_x_continuous(breaks=c(0,1,2),
                       labels=c(0,"l","2l"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0,1,2, irs_prod(2)),
                     labels=c(0,"f(l)", "f(2l)", "2f(l)"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Input, Labor (l)",
       y = "Output (q)",
       title = "Increasing Returns to Scale")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

```{r, fig.retina=3, fig.height=5.5}
irs_cost=function(x){x^{0.5}}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=irs_cost, geom="line", size=2, color="red")+
  #geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y = "Total Costs C(q)",
       title = "Total Costs with Increasing RTS")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

```{r, fig.retina=3, fig.height=5.5}
irs_mc=function(x){1/x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=irs_mc, geom="line", size=2, color="red")+
  geom_label(aes(x=2.5,y=irs_mc(2.5)), color = "red", label="MC(q)", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs per Unit",
       title = "Per Unit Costs with Increasing RTS")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```


## Cobb-Douglas Cost Functions

The total cost function for Cobb-Douglas production functions of the form
$$q=l^{\alpha}k^{\beta}$$
can be shown with some very tedious algebra to be:

$$C(w,r,q)=\left[\left(\frac{\alpha}{\beta}\right)^{\frac{\beta}{\alpha+\beta}} + \left(\frac{\alpha}{\beta}\right)^{\frac{-\alpha}{\alpha+\beta}}\right] w^{\frac{\alpha}{\alpha+\beta}} r^{\frac{\beta}{\alpha+\beta}} q^{\frac{1}{\alpha+\beta}}$$

If you take the first derivative of this (to get marginal cost), it is:

$$\frac{\partial C(w,r,q)}{\partial q}= MC(q) = \frac{1}{\alpha+\beta} \left(w^{\frac{\alpha}{\alpha+\beta}} r^{\frac{\beta}{\alpha+\beta}}\right) q^{\left(\frac{1}{\alpha+\beta}\right)-1}$$

How does marginal cost change with increased output? Take the second derivative:

$$\frac{\partial^2 C(w,r,q)}{\partial q^2}= \frac{1}{\alpha+\beta} \left(\frac{1}{\alpha+\beta} -1 \right) \left(w^{\frac{\alpha}{\alpha+\beta}} r^{\frac{\beta}{\alpha+\beta}}\right) q^{\left(\frac{1}{\alpha+\beta}\right)-2}$$

- If $\frac{1}{\alpha+\beta} > 1$, this is positive $\implies$ *decreasing returns to scale*
  - $\alpha+\beta < 1$ in the production function
- If $\frac{1}{\alpha+\beta} < 1$, this is negative $\implies$ *increasing returns to scale*
  - $\alpha+\beta > 1$ in the production function
- If $\frac{1}{\alpha+\beta} = 1$, this is constant $\implies$ *constant returns to scale*
  - $\alpha+\beta = 1$ in the production function

### Example (Constant Returns)

Let $q=l^{0.5}k^{0.5}$. 

$$\begin{align*}
C(w,r,q)&=\left[\left(\frac{0.5}{0.5}\right)^{\frac{0.5}{0.5+0.5}} + \left(\frac{0.5}{0.5}\right)^{\frac{-0.5}{0.5+0.5}}\right] w^{\frac{0.5}{0.5+0.5}} r^{\frac{0.5}{0.5+0.5}} q^{\frac{1}{0.5+0.5}}\\
C(w,r,q)&= \left[1^{0.5}+1^{-0.5} \right]w^{0.5}r^{0.5}q^{0.5}\\
C(w,r,q)&= w^{0.5}r^{0.5}q^{1}\\
\end{align*}$$

If $w=9$, $r=25$: 

$$\begin{align*}C(w=10,r=20,q)&=9^{0.5}25^{0.5}q \\ 
& =3*5*q\\
& =15q\\\end{align*}$$

```{r, fig.retina=3, fig.height=5.5}
cost_ex=function(x){15*x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=cost_ex, geom="line", size=2, color="red")+
  geom_label(aes(x=19,y=cost_ex(19)), color = "red", label="C(q)", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,400,50),
                     limits=c(0,400),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs C(q)",
       title = "Total Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

Marginal costs would be

$$MC(q) = \frac{\partial C(q)}{\partial q} = 15$$

Average costs would be

$$MC(q) = \frac{C(q)}{q} = \frac{15q}{q} = 15$$

```{r, fig.retina=3, fig.height=5.5}
mc_ex=function(x){15}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=mc_ex, geom="line", size=2, color="red")+
  geom_label(aes(x=17,y=15), color = "red", label="MC(q)=AC(q)", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Per Unit Costs",
       title = "Per Unit Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

### Example (Decreasing Returns)

Let $q=l^{0.25}k^{0.25}$. 

$$\begin{align*}
C(w,r,q)&=\left[\left(\frac{0.25}{0.25}\right)^{\frac{0.25}{0.25+0.25}} + \left(\frac{0.25}{0.25}\right)^{\frac{-0.25}{0.25+0.25}}\right] w^{\frac{0.25}{0.25+0.25}} r^{\frac{0.25}{0.25+0.25}} q^{\frac{1}{0.25+0.25}}\\
C(w,r,q)&= \left[1^{0.5}+1^{-0.5} \right]w^{0.5}r^{0.5}q^{2}\\
C(w,r,q)&= w^{0.5}r^{0.5}q^{2}\\
\end{align*}$$

If $w=9$, $r=25$: 

$$\begin{align*}C(w=10,r=20,q)&=9^{0.5}25^{0.5}q^2 \\ 
& =3*5*q^2\\
& =15q^2\\\end{align*}$$

```{r, fig.retina=3, fig.height=5.5}
cost_ex=function(x){15*x^2}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=cost_ex, geom="line", size=2, color="red")+
  geom_label(aes(x=19,y=cost_ex(19)), color = "red", label="C(q)", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,5000,500),
                     limits=c(0,5000),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs C(q)",
       title = "Total Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

Marginal costs would be

$$MC(q) = \frac{\partial C(q)}{\partial q} = 30q$$

Average costs would be

$$AC(q) = \frac{C(q)}{q} = \frac{15q^2}{q} = 15q$$

```{r, fig.retina=3, fig.height=5.5}
mc_ex_2=function(x){30*x}
ac_ex_2=function(x){15*x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=mc_ex_2, geom="line", xlim=c(0,20), size=2, color="red")+
  stat_function(fun=ac_ex_2, geom="line", xlim=c(0,20), size=2, color="orange")+
  geom_label(aes(x=8,y=mc_ex_2(8)), color = "red", label="MC(q)", size = 5)+
  geom_label(aes(x=18,y=ac_ex_2(18)), color = "orange", label="AC(q)", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,300,50),
                     limits=c(0,300),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Per Unit Costs",
       title = "Per Unit Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

