---
title: "Problem Set 2"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

*Due by Thursday, September 26, 2019* (Section 1 - TuTh)

*Due by Monday, September 30, 2019* (Section 2 - MW)

- [<i class="fas fa-file-pdf"></i> `Download as PDF`](/homeworks/02-problem-set-pdf.pdf)

- [<i class="far fa-life-ring"></i> `Answer Key`](/homeworks/02-problem-set-answers-pdf.pdf)

Please write your answers to the following questions on a piece of paper, or download and print a PDF copy (link above) to write on. You may also type your answers and print out a hard copy.

You may work together (and I highly encourage that) but you must turn in your own answers. Your TA, under my supervision, will grade homeworks 70% for completion, and for the remaining 30%, pick one question to grade for accuracy - so it is best that you try every problem, even if you are unsure how to complete it accurately.

# Concepts and Critical Thinking

Please answer the following questions briefly (1-3 sentences). Use examples as necessary. Be sure to label graphs fully, if appropriate.

## Question 1

Describe, in your own words, the (i) price effect, (ii) real income effect, and (iii) substitution effect from a price change.

## Question 2

Under what conditions can the law of demand be violated (however theoretical)?

## Question 3

For each of the following pairs, which of the two goods is more likely to have a *low* price elasticity of demand (less elastic) and why? 

### Part A
Demand for tangerines vs. demand for fruit

### Part B
Demand for beef next month vs. demand for beef over the next decade

### Part C
Demand for Exxon gasoline at the corner of 7th and Grand vs. demand for gasoline in the entire city

### Part D
Demand for insulin vs. demand for vitamins

## Question 4

Suppose that, holding prices constant, Alice has preferences over the number of books she purchases, illustrated in the table below.

```{r, echo = F, message = F, warning = F}
engel<-tribble(
  ~Income, ~Books,
  "$5", 5,
  "$10", 6,
  "$15", 20,
  "$20", 25,
  "$25", 26,
  "$30", 10,
  "$35", 9,
  "$40", 8,
  "$45", 7,
  "$50", 6
)

engel %>%
  knitr::kable(., format="html")
```

Draw a smooth approximation of Alice's Engel curve for books, and indicate the range(s) of income over which books are *normal* and/or *inferior* goods.

# Quantitative Applications

Show all work for calculations. You may lose points, even if correct, for missing work. Be sure to label graphs fully, if appropriate.

## Question 5

Steve spends his disposable income on meals at restaurants $(r)$ and paperback novels $(n)$. His usual restaurant meal costs $25, and paperback books cost $8. When Steve’s monthly income is $240, he goes out to eat 8 times and purchases 5 books. When his income rises to $282, he goes out to eat 10 times and purchases 4 books. 

### Part A

Calculate the income elasticity for meals at restaurants $(r)$. Is this an inferior, necessity, or luxury good?

### Part B

Calculate the income elasticity for paperback novels $(n)$. Is this an inferior, necessity, or luxury good?

## Question 6

Kendra buys eggs $(e)$, bagels $(b)$, and coffee $(c)$ for breakfast for the week.

### Part A

When eggs are $2/carton, she buys 5 bagels. When the price of eggs falls to $1/carton, she buys 4 bagels. Calculate the cross-price elasticity between eggs and bagels. Are they complements or substitutes for Kendra?

### Part B

When eggs are $2/carton, she buys 3 cups of coffee. When the price of eggs falls to $1/carton, she buys 6 cups of coffee. Calculate the cross-price elasticity between eggs and coffee. Are they complements or substitutes for Kendra?

## Question 7

Sketch a graph showing a *decrease* in the price of a good (on the horizontal axis, e.g. $x$ if you want). Indicate the (real) income effect, substitution effect, and price effect on the graph. Labelling points and describing each effect as a movement between specific points is sufficient. 

## Question 8

The demand for gym memberships is given by
$$q_D=500-5p$$

### Part A

Write the inverse demand function.

### Part B

Calculate the price elasticity of demand at a price of $80. Is this relatively elastic or relatively inelastic?

### Part C

What is the total revenue at a price of $80?

### Part D

Calculate the price elasticity of demand at a price of $10. Is this relatively elastic or relatively inelastic?

### Part E

What is the total revenue at $10?
 
### Part F

At what price is demand unit elastic, i.e. $\epsilon_D=-1$?

### Part G

What is the total revenue at the price you find in part (f)?